The resurgence of the linear optics quantum interferometer --- recent advances \& applications

### 2018-03-28 paper.tex
Final touch up
Added more references, cleaned up some citations, revised a few
grammatical errors

The promises offered by linear optics have spearheaded many \sihui{scientific and} technical advancements. One of the major successes for linear optics is \sihui{the detection of gravitational waves by the LIGO experiments based on the Michelson interferometer \cite{bib:LIGO2015}. Another is} the paradigm shift from bulk optics to integrated photonics \cite{bib:Meany15}. The advantage of using such integrated photonics over bulk optics is that it is more stable against phase fluctuations, and miniaturized. This has increased the scalability of optical implementations of quantum information protocols. Furthermore, it is now possible to have single-photon sources and detectors together with linear-optical networks on a silica chip \cite{bib:Sprengers11, bib:Silverstone14}. Having all components on chip reduces coupling losses which would be crucial for fault-tolerance \cite{bib:Li15}.

@@

Additional quantum labels are added to the creation and annihilation operators when other degrees of freedom, such as polarization, orbital angular momentum, and time-bins, are present. In this case, their commutator relation is
\begin{align}
[\ah_{j,\alpha},\adagh_{k,\beta}]=\delta_{j,k}\delta_{\alpha,\beta}\ ,
\end{align}
where $\alpha$ and $\beta$ represent these other degrees of freedom. As a consequence, quantum interference between multiple photons only occur when all quantum labels are the same. It is also possible to derive an analogous decomposition to that of Reck {\it et al.~}that realizes the unitary transformation on such photons as a sequence of beamsplitters and internal transformations.

@@

An alternative method to process tomography is feeding coherent states to probe the interferometer \cite{bib:Rahimi-Keshari13,bib:Heilmann15} instead of Fock states. Such states are produced by a standard laser source, thus reducing experimental resources. The $r_{jk}$ terms can be calculated from the ratios of output intensities at the $k$th port to the input intensity at the $j$th port. The remaining phases $\theta_{ij}$ are found by the interference pattern given by a two-mode coherent state $\ket{\alpha}\ket{\alpha e^{\phi}}$. When the states are injected into ports 1, and $j$ respectively, the output intensity at the $k$th port is
\begin{align}
I_k=I(r_{1k}^2+r_{jk}^2+2 r_{1k}r_{kj}\cos(\phi+\theta_{jk})) \ ,
\end{align}
where $\theta_{jk}=0$ for $k=1$, and $I$ is the intensity of the input coherent states. By scanning the phase shift $\phi$ and locating the maximum value of $I_k$ for $j=2,\ldots, m$, all unknown phases can be found via $\theta_{jk}=2\pi-\phi$. An elegant modification of this scheme removes the need for precise control of the phase-shift $\phi$ by suggesting instead to plot the output intensity $I_k$ with respect to the input intensity $I$ \cite{bib:Heilmann15}. In time, the natural drift in the laser source will cause this plot to trace out an ellipse, known as a Lissajous figure, whose orientation and direction of evolution will give the phase $\theta_{jk}$ and its sign respectively.

@@

Cluster states are a resource state for a model of quantum computation known as measurement-based quantum computation (MBQC). Here, having such a state as a resource enables universal quantum computation using only single-qubit measurements \cite{bib:Raussendorf03}. In fact, only (X,Y)-plane measurements are needed \cite{bib:Mantri17}. Therefore, the preparation of such states is highly valuable, generating much interest in their efficient preparation.

@@

\sihui{This concept for switching is widely used in current laboratories, however it is very slow using present-day technology. Although switching on its own can be completed in a timespan of the order of picoseconds, the process is slowed by the response time needed for its electronic control. Overcoming this hurdle is especially crucial in switching for fast-feedforward that is required, for example, in universal optical quantum computing. In this application, we would require total switching times, including time for control electronics to respond, on the order of nanoseconds, {\it i.e.~}faster than the time to the destination of the light it needs to switch, and this is still not possible. Presently this remains a daunting challenge, requiring much further research, with quantum memories being a candidate for loosening the switching time limitations.
}

@@

\subsection{Linear optical quantum computation (LOQC)}\label{sec:LOQC}
The advancements in linear optics have opened up avenues for implementation. One of the key requirements is nonlinear couplings between the different optical modes. Photons do not naturally interact with one another so some careful engineering is needed to achieve this coupling. Methods used include measurements and feedforwarding \cite{bib:KLM01}, and photon-atom interactions \cite{bib:Brod16}. The former requires non-deterministic operations, which succeed only part of the time and have to be repeated. This can be challenging given the limited resources to begin with. \sihui{Various linear-optical implementations of two-qubit gates important for approximately universal quantum computation are listed in Ref.~\cite{bib:Bartkowiak2010}, and comprehensive reviews have been written previously to cover this topic \cite{bib:JW12,bib:Kok05}, hence we will focus on recent advances.}

 Integrated photonic circuits has greatly improved the feasibility of LOQC. Their small size and interferometric stability enable the requisite quantum interference to happen in certain LOQC settings. Programmable circuits that are reusable have also emerged \cite{bib:Metcalf14,bib:Carolan15}. This saves resources and lead time for producing new circuits. Combining these chips with existing higher-efficiency sources and detectors will expand their capabilities. Loss can be mitigated by moving all components, {\it i.e.~}sources and detectors, onto the same chip as the linear optical circuit \cite{bib:Sprengers11, bib:Silverstone14}.  What remains is to have fast feedforward also on chip. This might pose the greatest challenge yet for universal LOQC. The bottleneck is in the speed of modulation in the feedforward circuitry; only a handful of experiments have been able to implement adaptive feedforward for the purposes of LOQC \cite{bib:Prevedel07,bib:Xiao-Song12,bib:Mikova12,bib:Zhao14}.

@@

  Although there has been tremendous progress in linear optics, it seems likely that an efficient, and fault-tolerant universal quantum computer will require a hybrid system. \sihui{Linear optics is still interesting in its own right for other tasks as discussed in the remainder of this sections. We also briefly mention here that single photons and waveplates are used in an already commercial application, namely, quantum key distribution (QKD). While QKD fundamentally does not require interferometry \cite{bib:BennetBrassard84,bib:Ekert91}, some QKD protocols rely on the Mach-Zehnder \cite{Bennett92, bib:Hughes2000,bib:Sasaki14,bib:Takesue15} and Franson interferometers \cite{bib:Ali-Khan07,bib:Brougham13}. Practical quantum key distribution relying on photonic technology is reviewed in \cite{bib:Scarani09,bib:Hoi-Kwong14}, and incredibly has been implemented at a distance of 1200 km by Earth-satellite transmissions \cite{bib:Liao17}. On the other hand, optimal quantum-optical cloning implementations \cite{bib:Fan2014} based on Mach-Zehnder interferometers have been applied to perform attacks on known quantum cryptographic schemes \cite{bib:Bartkiewicz2013,bib:Bartkiewicz2017}. Thus, linear optics remains fascinating platform for applications and foundation work. 
}

@@

The difficulty in producing nonlinear operations required for universal LOQC led researchers to seek out alternative ways for demonstrating its power in the near term. {\sc Bosonsampling} is the result of such an effort. It is a restricted model of non-universal quantum computation \cite{bib:Aaronson11} that constitutes sampling the photocounts from the probability distribution of identical bosons scattered by a passive linear optical interferometer. While not universal, the {\sc bosonsampling} scheme is strongly believed to implement a classically hard task. Consider such a circuit of $m$ input (and output) modes that is injected with $n$ indistinguishable single photons such that $m\sim \mathcal{O}(n^2)$. Then, the boson-sampling task consists of generating a sample from the probability distribution of single-photon measurements at the output of the circuit. The probability of detecting $s_j$ photons at the $j$th output mode is
\begin{align}
p(s_1,\ldots,s_m)=\frac{|{\rm Per}(U_S)|^2}{s_1!\ldots s_n!} \ ,	
\end{align}
%<<<<<<< HEAD:paper.tex
where $\text{Per}()$ denotes the permanent of a matrix, $S=\{s_1,\dots,s_m\}$ denotes an output photon-number configuration, and $U_S$ is obtained from the interferometer unitary $U$ by keeping its first $n$ columns and repeating its $j$th row $s_j$ times. The appearance of the permanent, which is \#P-hard in general (a complexity class even harder than NP-hard), in this statistics contributes to the hardness of the boson-sampling problem. Note that the number of configurations $|S|=\binom{n+m-1}{n}$ in the output superposition scales exponentially with $m$. For this reason, the boson sampler does not let us \emph{calculate} matrix permanents, as this would require knowing individual amplitudes with high precision, which would require an exponential number of samples. Rather, the boson sampler samples across a distribution of permanents without actually revealing any of them to the experimenter. Nonetheless, this is a classically intractable problem. It is also known that boson-sampling is as hard as stimulating the short time evolution of a certain Hamiltonian \cite{bib:Peropadre17}.

@@

As experimental imperfections threaten the scalability of {\sc bosonsampling}, there have been efforts to look into its error tolerance \cite{bib:Rohde12_tol}, fault-tolerance \cite{bib:Leverrier15}, and experimental artefacts \cite{bib:Shchesnobich14,bib:Tamma16}. Some works have resulted in interesting connections to exotic mathematical entities like immanants \cite{bib:Tan2013,bib:deGuise2014} and the multi-dimensional permanent \cite{bib:Tichy15}. A time-bin implementation of {\sc bosonsampling} \cite{bib:Motes14,bib:HeDing17} seems particularly attractive in terms of robustness \cite{bib:Motes15_timebin}.

@@

 As discussed in Section \ref{sec:bosonsampling}, the resources needed for Boson-sampling are readily available, but executes a task that is believed to be hard classically. Surprisingly, an entanglement between a large number of path labels is possible with single photons in a passive linear optical device. Motes {\it et al.~}harnessed this so-called number-path entanglement for quantum metrology without the use of entangled states \cite{bib:Motes15_LOQM}. Here, the phase shift to be estimated, $\varphi$, manifests itself as a linear phase gradient across $N$ modes (See Fig. \ref{fig:MORDOR}). By straddling these modes among a quantum Fourier transform and its conjugate operation, it is possible to achieve $\Delta \varphi=\mathcal{O}(N^{-3/2})$ using a coincidence measurement with just a string of $N$ single photons, for up to at least $N=25$ modes.

>>>>

The promises offered by linear optics have spearheaded many \sihui{scientific and} technical advancements. One of the major successes for linear optics is \sihui{the detection of gravitational waves by the LIGO experiments based on the Michelson interferometer \cite{bib:LIGO2015}. The further application of squeezed light to the detectors promises to reduce quantum noise, which currently limits their performance \cite{bib:Barsotti18}. Another advancement is} the paradigm shift from bulk optics to integrated photonics \cite{bib:Meany15}. The advantage of using such integrated photonics over bulk optics is that it is more stable against phase fluctuations, and miniaturized. This has increased the scalability of optical implementations of quantum information protocols. Furthermore, it is now possible to have single-photon sources and detectors together with linear-optical networks on a silica chip \cite{bib:Sprengers11, bib:Silverstone14}. Having all components on chip reduces coupling losses which would be crucial for fault-tolerance \cite{bib:Li15}.

@@

Additional quantum labels are added to the creation and annihilation operators when other degrees of freedom, such as polarization, orbital angular momentum, and time-bins, are present. In this case, their commutator relation is
\begin{align}
[\ah_{j,\alpha},\adagh_{k,\beta}]=\delta_{j,k}\delta_{\alpha,\beta}\ ,
\end{align}
where $\alpha$ and $\beta$ represent these other degrees of freedom. As a consequence, quantum interference between multiple photons only occur when all quantum labels are the same. It is also possible to derive an analogous decomposition to that of Reck {\it et al.~}that realizes the unitary transformation on such photons as a sequence of beamsplitters and internal transformations \cite{bib:Dhand2015}.

@@

An alternative method to process tomography is feeding coherent states to probe the interferometer \cite{bib:Rahimi-Keshari13,bib:Heilmann15} instead of Fock states. Such states are produced by a standard laser source, thus reducing experimental resources. The $r_{jk}$ terms can be calculated from the ratios of output intensities at the $k$th port to the input intensity at the $j$th port. The remaining phases $\theta_{ij}$ are found by the interference pattern given by a two-mode coherent state $\ket{\alpha}\ket{\alpha e^{i\phi}}$. When the states are injected into ports 1, and $j$ respectively, the output intensity at the $k$th port is
\begin{align}
I_k=I(r_{1k}^2+r_{jk}^2+2 r_{1k}r_{kj}\cos(\phi+\theta_{jk})) \ ,
\end{align}
where $\theta_{jk}=0$ for $k=1$, and $I$ is the intensity of the input coherent states. By scanning the phase shift $\phi$ and locating the maximum value of $I_k$ for $j=2,\ldots, m$, all unknown phases can be found via $\theta_{jk}=2\pi-\phi$. An elegant modification of this scheme removes the need for precise control of the phase-shift $\phi$ by suggesting instead to plot the output intensity $I_k$ with respect to the input intensity $I$ \cite{bib:Heilmann15}. In time, the natural drift in the laser source will cause this plot to trace out an ellipse, known as a Lissajous figure, whose orientation and direction of evolution will give the phase $\theta_{jk}$ and its sign respectively.

@@

Cluster states are a resource state for a model of quantum computation known as measurement-based quantum computation (MBQC). Here, having such a state as a resource enables universal quantum computation using only single-qubit measurements \cite{bib:Raussendorf03}. In fact, only $(X,Y)$-plane measurements are needed \cite{bib:Mantri17}. Therefore, the preparation of such states is highly valuable, generating much interest in their efficient preparation.

@@

\sihui{This concept for switching is widely used in current laboratories, however, it is very slow using present-day technology. Although switching on its own can be completed in a timespan of the order of picoseconds, the process is slowed by the response time needed for its electronic control. Overcoming this hurdle is especially crucial in switching for fast-feedforward that is required, for example, in universal optical quantum computing. In this application, we would require total switching times, including time for control electronics to respond, on the order of nanoseconds, {\it i.e.~}faster than the time to the destination of the light it needs to switch, and this is still not possible. Presently this remains a daunting challenge, requiring much further research, with quantum memories being a candidate for loosening the switching time limitations.
}

@@

\subsection{Linear optical quantum computation (LOQC)}\label{sec:LOQC}
The advancements in linear optics have opened up avenues for implementation. One of the key requirements is nonlinear couplings between the different optical modes. Photons do not naturally interact with one another so some careful engineering is needed to achieve this coupling. Methods used include measurements and feedforwarding \cite{bib:KLM01}, and photon-atom interactions \cite{bib:Brod16}. The former requires non-deterministic operations, which succeed only part of the time and have to be repeated. This can be challenging given the limited resources to begin with. \sihui{Various linear-optical implementations of two-qubit gates important for approximately universal quantum computation are listed in Ref.~\cite{bib:Bartkowiak2010}, and comprehensive reviews have been written previously to cover this topic \cite{bib:JW12,bib:Kok05, bib:Flamini2018}, hence we will focus on recent advances.}

 Integrated photonic circuits have greatly improved the feasibility of LOQC. Their small size and interferometric stability enable the requisite quantum interference to happen in certain LOQC settings. Programmable circuits that are reusable have also emerged \cite{bib:Metcalf14,bib:Carolan15}. This saves resources and lead time for producing new circuits. Combining these chips with existing higher-efficiency sources and detectors will expand their capabilities. Loss can be mitigated by moving all components, {\it i.e.~}sources and detectors, onto the same chip as the linear optical circuit \cite{bib:Sprengers11, bib:Silverstone14}.  What remains is to have fast feedforward also on chip. This might pose the greatest challenge yet for universal LOQC. The bottleneck is in the speed of modulation in the feedforward circuitry; only a handful of experiments have been able to implement adaptive feedforward for the purposes of LOQC \cite{bib:Prevedel07,bib:Xiao-Song12,bib:Mikova12,bib:Zhao14}.

@@

  Although there has been tremendous progress in linear optics, it seems likely that an efficient, and fault-tolerant universal quantum computer will require a hybrid system. \sihui{Linear optics is still interesting in its own right for other tasks as discussed in the remainder of this section. We also briefly mention here that single photons and waveplates are used in an already commercial application, namely, quantum key distribution (QKD). While QKD fundamentally does not require interferometry \cite{bib:BennetBrassard84,bib:Ekert91}, some QKD protocols rely on the Mach-Zehnder \cite{Bennett92, bib:Hughes2000,bib:Sasaki14,bib:Takesue15} and Franson interferometers \cite{bib:Ali-Khan07,bib:Brougham13}. Practical quantum key distribution relying on photonic technology is reviewed in \cite{bib:Scarani09,bib:Hoi-Kwong14}, and incredibly has been implemented at a distance of 1200 km by Earth-satellite transmissions \cite{bib:Liao17}. On the other hand, optimal quantum-optical cloning implementations \cite{bib:Fan2014} based on Mach-Zehnder interferometers have been applied to perform attacks on known quantum cryptographic schemes \cite{bib:Bartkiewicz2013,bib:Bartkiewicz2017}. Thus, linear optics remains fascinating platform for applications and foundation work. 
}

@@

The difficulty in producing nonlinear operations required for universal LOQC led researchers to seek out alternative ways for demonstrating its power in the near term. {\sc Bosonsampling} is the result of such an effort. It is a restricted model of non-universal quantum computation \cite{bib:Aaronson11} that constitutes sampling the photocounts from the probability distribution of identical bosons scattered by a passive linear optical interferometer. While not universal, the {\sc bosonsampling} scheme is strongly believed to implement a classically hard task. Consider such a circuit of $m$ input (and output) modes that is injected with $n$ indistinguishable single photons such that $m\sim \mathcal{O}(n^2)$. Then, the boson-sampling task consists of generating a sample from the probability distribution of single-photon measurements at the output of the circuit. The probability of detecting $s_j$ photons at the $j$th output mode is
\begin{align}
p(s_1,\ldots,s_m)=\frac{|{\rm Per}(U_S)|^2}{s_1!\ldots s_n!} \ ,	
\end{align}
%<<<<<<< HEAD:paper.tex
where $\text{Per}()$ denotes the permanent of a matrix, $S=\{s_1,\dots,s_m\}$ denotes an output photon-number configuration, and $U_S$ is obtained from the interferometer unitary $U$ by keeping its first $n$ columns and repeating its $j$th row $s_j$ times. The appearance of the permanent, which is \#P-hard in general (a complexity class even harder than NP-hard) \sihui{to compute}, in this statistics contributes to the hardness of the boson-sampling problem. Note that the number of configurations $|S|=\binom{n+m-1}{n}$ in the output superposition scales exponentially with $m$. For this reason, the boson sampler does not let us \emph{calculate} matrix permanents, as this would require knowing individual amplitudes with high precision, which would require an exponential number of samples. Rather, the boson sampler samples across a distribution of permanents without actually revealing any of them to the experimenter. Nonetheless, this is a classically intractable problem. It is also known that {\sc BosonSampling} is as hard as stimulating the short time evolution of a certain Hamiltonian \cite{bib:Peropadre17}.

@@

As experimental imperfections threaten the scalability of {\sc bosonsampling}, there have been efforts to look into its error tolerance \cite{bib:Rohde12_tol}, fault-tolerance \cite{bib:Leverrier15}, and experimental artefacts \cite{bib:Shchesnobich14,bib:Tamma16}. Some works have resulted in interesting connections to exotic mathematical entities like immanants \cite{bib:Tan2013,bib:deGuise2014} and the multi-dimensional permanent \cite{bib:Tichy15}. A time-bin implementation of {\sc bosonsampling} \cite{bib:Motes14,bib:HeDing17} seems particularly attractive in terms of robustness \cite{bib:Motes15_timebin}. \sihui{More discussions of {\sc BosonSampling} can be found in the review paper that is Ref.~\cite{bib:Lund2017}}.

@@

 As discussed in Section \ref{sec:bosonsampling}, the resources needed for Boson-sampling are readily available, but executes a task that is believed to be hard classically. Surprisingly, an entanglement between a large number of path labels is possible with single photons in a passive linear optical device. Motes {\it et al.~}harnessed this so-called number-path entanglement for quantum metrology without the use of entangled states \cite{bib:Motes15_LOQM,bib:Olsen2017}. Here, the phase shift to be estimated, $\varphi$, manifests itself as a linear phase gradient across $N$ modes (See Fig.~\ref{fig:MORDOR}). By straddling these modes among a quantum Fourier transform and its conjugate operation, it is possible to achieve $\Delta \varphi=\mathcal{O}(N^{-3/2})$ using a coincidence measurement with just a string of $N$ single photons, for up to at least $N=25$ modes. \sihui{Shortly after its construction, this scheme was realized experimentally \cite{bib:Zu-En2017}. The quantum Cram\'{e}r-Rao bound of estimating $\varphi$ in such interferometers was also analytically computed \cite{bib:Chenglong2017}.}
